---
title: "Week 1 Lab"
format: pdf
editor: visual
---

## Lab Exercises

```{r}
#install.packages("tidyverse")
library(tidyverse)
```

```{r}
dm <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Mx_1x1.txt", skip = 2, col_types = "dcddd")
head(dm)
```

1.  Plot the ratio of male to female mortality rates over time for ages 10,20,30 and 40 (different color for each age) and change the theme

```{r}
dlab <- dm |>  
  mutate(mf_ratio = Male/Female) |>
  filter(Age==10|Age==20|Age==30|Age==40) |>
  select(Year:mf_ratio)

dlab |>
  ggplot(aes(x = Year, y = mf_ratio, color = Age)) +
  geom_line() +
  theme_minimal()+
  labs(title="Male:Female Mortality")
```

2.  Find the age that has the highest female mortality rate each year

```{r}
dm |>
  group_by(Year) |>
  filter(Female==max(Female, na.rm=TRUE)) |>
  select(Year, Age)

```

3.  Use the `summarize(across())` syntax to calculate the standard deviation of mortality rates by age for the Male, Female and Total populations.

```{r}
dm |> 
  group_by(Age) |> 
  summarize(across(c("Male","Female","Total"),sd,na.rm=TRUE))

```

4.  The Canadian HMD also provides population sizes over time (https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt). Use these to calculate the population weighted average mortality rate separately for males and females, for every year. Make a nice line plot showing the result (with meaningful labels/titles) and briefly comment on what you see (1 sentence). Hint: `left_join` will probably be useful here. Reformat existing dataset:

```{r}
dm4 <- dm |> 
select(Year:Male)
```

Get new dataset:

```{r}
df <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt", skip = 2, col_types = "dcddd")
```

Renaming Columns:

```{r}
df=rename(df,pop_f=Female,pop_m=Male)
```

Combine Datasets:

```{r}
df2 <- dm4 |> 
  left_join(df)
df2
```

Add Total Deaths by Sex:

```{r}
df3 <- df2 |>
  mutate(deaths_f=Female*pop_f, deaths_m=Male*pop_m)
```

Group by Year:

```{r}
df4 <- df3 |>
  group_by(Year) |>
  summarise(across(pop_f:deaths_m,sum,na.rm=TRUE))
```

Calculate Population-weighted Mortality by Year:

```{r}
df5 <- df4 |>
  mutate(Female=deaths_f/pop_f,Male=deaths_m/pop_m) |>
  pivot_longer(Female:Male,names_to="Sex",values_to="Weighted_Mortality")
```

Plot the Chart:

```{r}
df5 |>
  ggplot(aes(x=Year,y=Weighted_Mortality,color=Sex))+
  geom_line()+
  labs(title="Population-Weighted Average Mortality",
       y="Average Mortality Rate (Population-Weighted)")
```

The population-weighted average mortality for both sexes trended downwards until the early 2000s when it started to trend upwards; the male average is significantly higher than the female average, though also since the early 2000s, the two are becoming much closer aligned.
